The Two Dualities of Computation: Negative and Fractional Types

### 2012-02-28 rational.tex

\usepackage{graphicx}
%%\usepackage{longtable}
%%\usepackage{comment}
\usepackage{amsmath}
%%\usepackage{mdwlist}
%%\usepackage{txfonts}
\usepackage{xspace}
%%\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{multicol}
%%\usepackage[nodayofweek]{datetime}
%%\usepackage{etex}
%%\usepackage[all, cmtip]{xy}
%%\usepackage{xcolor}
%%\usepackage{listings}
%%\usepackage{bm}
\usepackage{cmll}

@@

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! eta = \eta
%! eps = \epsilon
%! eta+ = \eta^+
%! eps+ = \epsilon^+
%! eta* = \eta^*
%! ^^^ = ^{-1}
%! eps* = \epsilon^*
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +

@@

We show a deep symmetry between functions and delimited continuations, values
and continuations that arises in {{langRev}} in a manner that is reminiscent
of Filinksi's Symmetric \lcal ~\cite{Filinski:1989:DCI:648332.755574}. The
symmetry arises by extending {{langRev}} with a notion of additive duality
over the monoid {{(+, 0)}} by including {{eta}} and {{eps}} operators of
Compact Closed Categories. The resulting dual types, which we denote {{-b}},
have a time traveling ``backward information flow'' interpretation and allow
for the encoding of higher-order function and iteration via the construction
of trace operators, thereby making the extended language {{langRevEE}} a
Turing-complete reversible programming language with higher-order functions
and first-class delimited continuations.

>>>>

\usepackage{graphicx}
%%\usepackage{longtable}
%%\usepackage{comment}
\usepackage{amsmath}
%%\usepackage{mdwlist}
%%\usepackage{txfonts}
\usepackage{xspace}
%%\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{multicol}
\usepackage[nodayofweek]{datetime}
\usepackage{etex}
\usepackage[all, cmtip]{xy}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{multicol}
\newcommand\hmmax{0} % default 3
\newcommand\bmmax{0} % default 4
\usepackage{bm}
\usepackage{cmll}

@@

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! eta = \eta
%! eps = \epsilon
%! eta+ = \eta^+
%! eps+ = \epsilon^+
%! eta* = \eta^{\times}
%! eps* = \epsilon^{\times}
%! trace+ = trace^+
%! trace* = trace^{\times}
%! ^^^ = ^{-1}
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +

@@

We show a deep symmetry between functions and delimited continuations, values
and continuations that arises in {{langRev}} in a manner that is reminiscent
of Filinksi's Symmetric \lcal ~\cite{Filinski:1989:DCI:648332.755574}. The
symmetry arises by extending {{langRev}} with a notion of additive duality
over the monoid {{(+, 0)}} by including {{eta+}} and {{eps+}} operators of
Compact Closed Categories. The resulting dual types, which we denote {{-b}},
have a time traveling ``backward information flow'' interpretation and allow
for the encoding of higher-order function and iteration via the construction
of trace operators, thereby making the extended language {{langRevEE}} a
Turing-complete reversible programming language with higher-order functions
and first-class delimited continuations.
